DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life

### 2022-01-14 manuscript/manuscript.Rmd
updating description

Up to the time this manuscript was written, `datelife`'s chronogram database had `r length(opentree_chronograms$trees)` chronograms, pulled entirely from OpenTree's tree repository, the only public tree repository from where `datelife` can currently get chronograms to construct its database. This represents `r round(length(opentree_chronograms$trees)*100/3987, digits = 2)`% of the largest existing chronogram database, TimeTree, which has a collection of 3,998 chronograms as of `r format(Sys.Date(), format="%B %d %Y")`. Unfortunately, TimeTree's database is not open for scientific reuse nor automated data mining [@Kumar2017].
In 2015, a synthetic chronogram was constructed from 2,274 chronograms available at the time on the TimeTree database [@Hedges2015]. This is the only synthetic TimeTree chronogram that has been made publicly available and deposited on the OpenTree repository, and it is part of datelife's database now. Hence, the amount of lineages represented in datelife's database (`r length(unique(unname(unlist(lapply(opentree_chronograms$trees, "[[", "tip.label")))))` unique terminal taxa), is at least as substantial as TimeTree's (as of the last released tree, with `r max(sapply(opentree_chronograms$trees[grepl("Hedges", names(opentree_chronograms$trees))], ape::Ntip))` taxa). Regrettably, this does not ensure that the full state of knowledge of time of divergence of the taxon/lineage will be available.
<!--
Even though datelifes chronograms cover this many species from the tree of life compared to this many covered by TimeTree (database or synthetic chronogram?)-->
Incorporation of more published chronograms into `datelife`'s database is crucial to improve its services. One option to increase our database is the Dryad data repository.  Methods to automatically mine chronograms from Dryad could be designed and implemented. However, Dryad's metadata system has no information to automatically detect branch length units, and those would still need to be determined on a second step, by a curator.
<!-- Can we mine journal databases? -->
<!-- At the same time, the prospect is to increase awereness and keep inviting and doing work encouraging scientists to make their data sets available in OpenTree's database. -->
Consequently, we would like to emphasize on the importance of sharing chronogram data for the benefit of the scientific community as a whole, into repositories that require expert input and manual curation, such as OpenTree's tree repository [@mctavish2015phylesystem].

>>>>

At the time of writing of this manuscript, `datelife`'s database has `r length(opentree_chronograms$trees)` chronograms, pulled entirely from OpenTree's repository, the only public tree repository from where one can currently get chronograms to construct a database. This represents `r round(length(opentree_chronograms$trees)*100/3987, digits = 2)`% of the largest existing chronogram database, TimeTree, which has a collection of 3,998 chronograms as of `r format(Sys.Date(), format="%B %d %Y")`. Unfortunately, TimeTree's database is not open for scientific reuse nor automated data mining [@Kumar2017].
In 2015, a synthetic chronogram was constructed from 2,274 chronograms available at the time on the TimeTree database [@Hedges2015]. This is the only synthetic TimeTree chronogram that has been made publicly available and deposited on the OpenTree repository, and it is part of datelife's database now. Hence, the amount of lineages represented in datelife's database (`r length(unique(unname(unlist(lapply(opentree_chronograms$trees, "[[", "tip.label")))))` unique terminal taxa), is at least as substantial as TimeTree's (as of the last released tree, with `r max(sapply(opentree_chronograms$trees[grepl("Hedges", names(opentree_chronograms$trees))], ape::Ntip))` taxa). Regrettably, this does not ensure that the full state of knowledge of time of divergence of the taxon/lineage will be available.
<!--
Even though datelifes chronograms cover this many species from the tree of life compared to this many covered by TimeTree (database or synthetic chronogram?)-->
Incorporation of more published chronograms into `datelife`'s database is crucial to improve its services. One option to increase our database is the Dryad data repository.  Methods to automatically mine chronograms from Dryad could be designed and implemented. However, Dryad's metadata system has no information to automatically detect branch length units, and those would still need to be determined on a second step, by a curator.
<!-- Can we mine journal databases? -->
<!-- At the same time, the prospect is to increase awereness and keep inviting and doing work encouraging scientists to make their data sets available in OpenTree's database. -->
Consequently, we would like to emphasize on the importance of sharing chronogram data for the benefit of the scientific community as a whole, into repositories that require expert input and manual curation, such as OpenTree's tree repository [@mctavish2015phylesystem].
